Exploration and analysis of the WHO mortality database.

This notebook walks us trough a typical workflow of data analysis. From tiding the data to appling a linear regression modell and in the end communicate the results of our analysis.
We dont want to show some kind of typical textbook example. Our goal is to display the real process. This means that we will ask questions about the data in our dataset and answer them with transformation and visualisation techniques. 
In the end we want to apply a feasible linear regression model in order to gain insight into the futre development of the data.

To do all this we use the WHO mortality database, which lists all kinds of possibile types of deaths in the WHO member states.
The members of the WHO report the data from their civil registration systems into the database annually. We use the ICD-9 version which ranges from 1979 to 2017.

This are the sources from the WHO.
https://www.who.int/data/data-collection-tools/who-mortality-database


Import
We import these different files into R. The countrycodes file connects the countrys number to the name. Morticd9 is the mortality dataset itself. The population file contains informations about the population divided into the different age groups.

```{r}
setwd("C:/Users/Jan/Documents/GitHub/DASC/resources")
countrycodes <- read.csv("country_codes")
mortalityicd9 <- read.csv("Morticd9")
population <- read.csv("pop")
```
Restricting the dataset

We want to compare communist and democratic countries during the cold war. Only countries that were either in the NATO or in the warsaw pact, until 1991 apply. Unfurtunatly our dataset is inconsistent because it does not ranges from the begining of Nato or the cold war. Therefore we only use data ranging from 1979 to 1991. 
The Nato countries that are part of our data set are:
USA, Canada, Portugal, Turkey, Greece, Italy, Austria, France, West Germany, Belgium,Luxembourg, Netherlands, Britan, Norway, Iceland. 

The warsaw pact countries in our dataset are:
Albania, Bulgaria, East Germany, Poland, Rumania, UDSSR, Czechoslovakia, Hungary.

Splitting the Dataset

We split the raw data and afterwards tidy it. This code is used to split the data.
```{r}
setwd("C:/Users/Jan/Documents/GitHub/DASC/resources")
dividedbyyears <- subset(mortalityicd9, Year > 1979 & Year < 1991)

countrys <- c("United States of America", "Canada", "Portugal", "Turkey", "Greece", "Germany, Former Democratic Republic", "Germany, Former Federal Republic", "	
Germany, West Berlin", "Austria", "France", "Belgium", "Luxembourg", "Netherlands", "United Kingdom", "Norway", "Iceland", "Albania", "Bulgaria", "Poland", "Rumania", "USSR, Former","Czechoslovakia, Former", "Hungary")

coldwarmortality <- subset(dividedbyyears,)
```


