Exploration and analysis of the WHO mortality database.

This notebook walks us trough a typical workflow of data analysis. From tiding the data to applying a linear regression model and in the end communicate the results of our analysis.
We do not want to show some kind of typical textbook example. Our goal is to display the real process. This means that we will ask questions about the data in our dataset and answer them with transformation and visualisation techniques. 
In the end we want to apply a feasible linear regression model in order to gain insight into the future development of the data.

To do all this we use the WHO mortality database, which lists all kinds of possible types of deaths in the WHO member states.
The members of the WHO report the data from their civil registration systems into the database annually. We use the ICD-9 version which ranges from 1979 to 2017.

This are the sources from the WHO.
https://www.who.int/data/data-collection-tools/who-mortality-database

Packages
All the needed packages and librarys are defined here.
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/Jan/Documents/GitHub/DASC/resources')
library(dplyr)
library(ggplot2)
```

Import
We import these different files into R. The countrycodes file connects the countries` number to the name. Morticd9 is the mortality dataset itself. The population file contains information about the population divided into the different age groups. All files except the detailed causes can be found in our source. In order to get the detailed causes as a .csv file we copied the table from the word file into excel and then exported it as csv.

```{r}
countrycodes <- read.csv("country_codes")
mortalityicd9 <- read.csv("Morticd9")
population <- read.csv("pop")
detailedcause <- read.csv(file = "ICD-9BTL.csv", head = TRUE, sep=";")
```

Tidying the dataset

The death cause categories are divided into two different types. The "Detailed List Numbers" which are numbers connected to a cause and the BTL (Basic Tabulation List). BTL categorizes the numbers, for example B24 contains the numbers 380-389. But BTL is further divided, so B240 is a subcategory of B24 containing the numbers 381-383.
In the dataset B24 also contains all deaths of its subcategories.The targeted audience of this document are no health experts. Therefore only the BTL categories with two numbers are included.
The columns Deaths2, Deaths3,... stand for age groups, but Deaths1 sums up all age groups and Infants.
Therefore all other death columns get excluded.
In the Documentation there is a chapter 17 for ICD-9 all these "Detailed List Numbers" are already included in the BTL categories. Therefore they get thrown out.

```{r}
dataset <- mortalityicd9 %>%
  left_join(countrycodes, by = c("Country" = "country"))%>%
  rename("Countryname" = "name") %>%
  subset(select = c(1, 4, 6, 7, 10, 40))%>%
  select(Countryname, Country, Year, Cause, Sex, Deaths1)%>%
  rename("Deathcount" = "Deaths1")%>%
  subset(grepl("^B\\d{0,2}$", Cause))

subcategories <- mortalityicd9 %>%
  left_join(countrycodes, by = c("Country" = "country"))%>%
  rename("Countryname" = "name") %>%
  subset(select = c(1, 4, 6, 7, 10, 40))%>%
  select(Countryname, Country, Year, Cause, Sex, Deaths1)%>%
  rename("Deathcount" = "Deaths1")%>%
  subset(!grepl("^B\\d{0,2}$", Cause))%>%
  subset(grepl("^B", Cause)) #I don't know why I had to do this twice in order to eliminate numbers
```

Restricting the dataset

We want to compare communist and democratic countries during the cold war. Only countries that were either in the NATO or in the Warsaw pact, until 1991 apply. Unfortunately our dataset is inconsistent because it does not ranges from the creation of NATO or the start of  the cold war. Therefore we only use data ranging from 1979 to 1991. 
The NATO countries that are part of our data set are:
USA, Canada, Portugal, Turkey, Greece, Italy, Austria, France, West Germany, Belgium,Luxembourg, Netherlands, Britain, Norway, Iceland. 

The Warsaw pact countries in our dataset are:
Albania, Bulgaria, East Germany, Poland, Romania, USSR, Czechoslovakia, Hungary.

Splitting the Dataset

We split the raw data and afterwards tidy it. This code is used to split the data.
```{r}
#List or Vector of countries have to check that
warsawpact <- c("Germany, Former Federal Republic", "Albania", "Bulgaria", "Poland", "Romania", "USSR, Former","Czechoslovakia, Former", "Hungary", "Russian Federation")

natostates <- c("United States of America", "Canada", "Portugal", "Turkey", "Greece", "Germany, Former Democratic Republic", "Germany, Former Federal Republic", "	
Germany, West Berlin", "Austria", "France", "Belgium", "Luxembourg", "Netherlands", "United Kingdom", "Norway", "Iceland")

warsawpactcountries <- data.frame(matrix(unlist(warsawpact), nrow=length(warsawpact), byrow=T))%>%
  rename("Countryname" = "matrix.unlist.warsawpact...nrow...length.warsawpact...byrow...T.")

natocountries <- data.frame(matrix(unlist(natostates), nrow=length(natostates), byrow=T))%>%
  rename("Countryname" = "matrix.unlist.natostates...nrow...length.natostates...byrow...T.")

countryslist <- c(warsawpact, natostates)

coldwarcountrys <- data.frame(matrix(unlist(countryslist), nrow=length(countryslist), byrow=T))%>%
  rename("Countryname" = "matrix.unlist.countryslist...nrow...length.countryslist...byrow...T.")

#The different dataset divided by affiliation
coldwardataset <- dataset %>%
  subset(Year >= 1979 & Year <= 1991) %>%
  semi_join(coldwarcountrys)

warsawpactdataset <- dataset%>% 
  subset(Year >= 1979 & Year <= 1991) %>%
  semi_join(warsawpactcountries)

natodataset <- dataset%>% 
  subset(Year >= 1979 & Year <= 1991) %>%
  semi_join(natocountries)

```


Question1
Is the suicide rate higher in countries part of the warsaw pact?
```{r}
suiciderate <- coldwardataset%>%
  filter(Cause == "B54")

#This pick function
pick <- function(condition){
  function(d) d %>% filter_(condition)
}

ggplot(suiciderate, aes(x = Year, y = Deathcount))+
  geom_point()

suicideranking <- suiciderate%>%
  group_by(Countryname)%>%
  summarise(Deathcount = sum(Deathcount))%>%
  arrange(desc(Deathcount))
  
```
We can see that the date seems to be quite off.  
```{r}
ggplot(suiciderate, aes(x = Year, y = Deathcount))+
  geom_line(data = pick(~Countryname == "USSR, Former" & Sex==2), colour = "pink")+
  geom_line(data = pick(~Countryname == "USSR, Former" & Sex==1), colour = "blue")

ggplot(suiciderate, aes(x = Year, y = Deathcount))+
  geom_line(data = pick(~Countryname == "United States of America" & Sex==1), colour = "blue")+
  geom_line(data = pick(~Countryname == "United States of America" & Sex==2), colour = "pink")

ggplot(suiciderate, aes(x = Year, y = Deathcount))+
  geom_line(data = pick(~Countryname == "Russian Federation" & Sex==1), colour = "blue")+
  geom_line(data = pick(~Countryname == "Russian Federation" & Sex==2), colour = "pink")
```
We can see that the suicide rate in the former USSR is way higher then anywhere else. 
We can see that the Russian Federation has the second highest overall suicide rate.
Its deaths are part of the USSR. The Russian Federation is equivalent to todays russia. 
Its a more rural area compared to the other states of the USSR.
The reason for the high suicide rates is the harmful impact of the Soviet regime on rural life. 
Collective farms were organized against peoples will. Rural areas lacked food supply, medical care and social welfare. 
Citate
The USSR shows a big drop in suicides in 1985. The reason for that is the Perestroika. 
Perestroika served as a positive social factor, reducing the number of suicides. It is hard to tell which
contributed more, psychological freedom and hopes
for the future or restrictions on the sale of alcohol. 
Citate
https://pubmed.ncbi.nlm.nih.gov/1414406/#:~:text=Abstract,1965%20to%2029.6%20in%201984.

```{r}
tofilter <- suicideranking%>%
  tail(-5)

suicidesmoothed <- suiciderate %>%
  semi_join(tofilter, by="Countryname")

ggplot(suicidesmoothed, aes(x = Year, y = Deathcount))+
  geom_smooth(data = pick(~Sex==1), color="Blue")+
  geom_smooth(data = pick(~Sex==2), color="Pink")
```
In order to better compare the different countries, the top five countries are exclouded.
After that I compared male and female suicide rates. The diagram shows that the male suicide rate is way
higher than the female one.


```{r}
ggplot(suicidesmoothed, aes(x = Year, y = Deathcount))+
  geom_smooth(data = pick(~Countryname %in% warsawpact), color="Red")+
  geom_smooth(data = pick(~Countryname %in% natostates), color="Blue")
```
When comparing the countries by membirship to either the warsaw pact or nato, 
we see that the suicide rate from countries in the warsaw pact are way higher. 
Even when the USSR and Russian Federation is excluded.

```{r}
test <- subcategories %>%
  subset(Year >= 1979 & Year <= 1991) %>%
  semi_join(coldwarcountrys)%>%
  filter(Cause=="B037")

test2 <- warsawpactdataset%>%
  filter(Cause=="B04")
  #filter(Countryname!="USSR, Former" & Countryname!="Russian Federation" & Countryname!="Germany, Former Federal Republic")
  
ggplot(test, aes(x=Year, y=Deathcount, group=Countryname, color=Countryname))+
  geom_point()
```


```{r}

  
poisoning <- coldwardataset%>%
  filter(Cause=="B48" | Cause=="B50")%>%
  filter(!Countryname=="USSR, Former")%>%
  filter(!Countryname=="Russian Federation")

ggplot(poisoning, aes(x=Year, y=Deathcount, group=Countryname, color=Countryname))+
  geom_point()
```


TODO
1. Add Zeros in cause.csv
  The zeros in the second column are missing reimport, rename and delete useless rows.
2. Add proper citations in question1
3. Switch from geom_point to linear
4. Explain semi join
5. Explain what you did during tiding
6. Explain suicideRanking
7. Clear naming convention 
   In the beginning divide into usefull clear datasets. Then divide by disease.
   Or switch to lesser datasets, when there are too much datasets. 
8. Description for every ggplot. Legende

